MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments

### 2021-10-26 main/MSstatsPTM_Main.tex
Removed duplicate info in supp, changed headers in main

\renewcommand{\deg}{\ensuremath{^{\circ}}\xspace}

@@

\author[1]{Devon~Kohler}
\author[2]{Tsung-Heng~Tsai}
\author[1]{Ting~Huang}
\author[4]{Erik~Verschueren}
\author[3]{Trent~Hinkle}
\author[3]{Meena~Choi*}
\author[1]{Olga~Vitek*}
\affil[1]{Khoury College of Computer Science, Northeastern University, Boston, MA, USA}
\affil[2]{Kent State University, Kent, OH, USA}
\affil[3]{MPL, Genentech, South San Francisco, CA, USA}
\affil[4]{Galapagos, Mechelen, Antwerp, Belgium}
\affil[*]{Corresponding Authors}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Statistical modeling and parameter estimation}

The proposed approach takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance\cite{Choi:2014}. This approach allows the method to be extended to cases where the experimental design in inbalanced, and where additional sources of variation are present. In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site\cite{Tukey:1977}. After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models\cite{Bolker2009} \cite{Faraway:2006}. Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Missing value imputation}
\label{sec:missing_imp}

$MSstatsPTM$ gives the ability to impute missing feature intensities if desired. When values are imputed, it is assumed they are missing for reasons of low abundance. Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model\cite{Tukey:1977}. In order to impute a feature's missing values, the feature must be present in at least one MS Run. If the feature is not present across all MS Runs, the value will be left missing. Missing value imputation is done before summarization with Tukey's Median Polish in order to correct for the impact of outliers. Missing value imputation is done separately for both the PTM and global protein datasets.

@@

The methods when applied to a balanced design with multiple conditions and one technical replicate. In practice the experimental design can be unbalanced and variation can come from multiple sources. We show how the methods can be applied to more complex experimental designs in the next section.

Detection of differentially modified PTM sites is performed through testing the null hypothesis of ‘no change’ against the alternative. The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance. Notationally, given two conditions $i$ and $i'$, the hypothesis test is formulated as follows:

\begin{align*}
H_{0}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &= 0 \\
H_{a}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &\neq 0
\end{align*}

@@

\[
\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right],
\]

@@

\[
SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
\]

Finally the degrees of freedom are given by:

\[
\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
\]

The test statistic for the hypothesis testing is the ratio of the estimate of the adjusted difference to its SE. To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the t distribution with degrees of freedom approximated by the Satterthwaite method\cite{satterthwaite:1946}. Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$\cite{Benjamini:1995}. More details on how the PTM model is adjusted are provided in Supplementary Sec. \ref{supp-sec:test}

@@

The statistical modeling approaches discussed above can also be extended to complex designs including experiments with additional sources of variation and unbalanced designs. The null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$. 
\begin{align*}
H_{0}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &= 0 \\
H_{a}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &\neq 0
\end{align*}

Model inference is done in the same way to how MSstats~\cite{choi_etal_14a} and MSstatsTMT~\cite{Huang:2020} target the global protein. In cases where the design in unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood is used to estimate the parameters of each model. Once the parameters are estimated, we can combine the models using modified versions of the formulas in Section \ref{sec:adjustment_summary}.

The log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
\[
\hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
\]
and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
\[
SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
\]
%\[
%\left[ \left( \frac{2}{J} \right) \cdot \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
%\]
The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
\[
\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
\]

@@

\paragraph{Code availability}\mbox{}\\

\noindent{{\bf Bioconductor:} http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html}

\noindent{{\bf Github:} https://github.com/Vitek-Lab/MSstatsPTM}

@@

\subsubsection*{Dataset 1 : Computer simulation 1 - Label-free}

@@

\subsubsection*{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}

@@

\subsubsection*{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}

@@

\subsubsection*{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}

@@

\subsubsection*{Sample size calculation}

Here we analyze the sample size needed to achieve a desired statistical power. The proposed approach corrects for confounding between the modified peptide and unmodified protein at a cost of increased variation. This can be seen in the calculation for variance in Section \ref{sec:adjustment_summary}. When the variation is increased, the experiment requires a larger number of replicates to reach the same power. Thus the statistical power is dependent on the variance from both the modified peptide and unmodifed protein as well as the number of replicates in both runs.

>>>>

\renewcommand{\deg}{\ensuremath{^{\circ}}\xspace}
\newcommand{\dd}[1]{\mathrm{d}#1}

@@

\author[1]{Devon~Kohler}
\author[2]{Tsung-Heng~Tsai}
\author[1]{Ting~Huang}
\author[4]{Erik~Verschueren}
\author[3]{Trent~Hinkle}
\author[3]{Meena~Choi*}
\author[1]{Olga~Vitek*}
\affil[1]{Khoury College of Computer Science, Northeastern University, Boston, MA, USA}
\affil[2]{Kent State University, Kent, OH, USA}
\affil[3]{MPL, Genentech, South San Francisco, CA, USA}
\affil[4]{ULUA BV, Arendstraat 29, 2018 Antwerp, Belgium}
\affil[*]{Corresponding Authors}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Existing work in statistical modeling and parameter estimation}

The proposed approach first leverages the large amount of work done to create MSstats~\cite{Choi:2014}. It takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance\cite{Choi:2014}. This approach allows the method to be extended to cases where the experimental design in inbalanced, and where additional sources of variation are present. In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site\cite{Tukey:1977}. After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models\cite{Bolker2009} \cite{Faraway:2006}. Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

Additionally, the proposed method can include imputing missing feature intensities if desired. When values are imputed, it is assumed they are missing for reasons of low abundance. Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model\cite{Tukey:1977}. In order to impute a feature's missing values, the feature must be present in at least one MS Run. If the feature is not present across all MS Runs, the value will be left missing. Missing value imputation is done before summarization with Tukey's Median Polish in order to correct for the impact of outliers. Missing value imputation is done separately for both the PTM and global protein datasets.

@@

Here we propose a statistical method for detecting changes in PTMs and removing confounding with changes in global protein abundance. The method is shown when applied to a balanced design with multiple conditions and one technical replicate. In practice the experimental design can be unbalanced and variation can come from multiple sources. We show how the methods can be applied to more complex experimental designs in the next section.

Detection of differentially modified PTM sites is performed through testing the null hypothesis of ‘no change’ against the alternative. The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance. There are $I$ conditions and $J$ mass spectrometry runs (technical replicates) per condition in the experiment. The PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. Conversely, the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$. The expected values of log-abundances of the PTM and protein in Condition $i$ are denoted by $\mu_{i}^{\ast}$ and $\mu_{i}$. The feature and run level summarization for condition $i$ is denoted as $\hat{y}_{i+}^{\ast}$ for the modified peptide and $\hat{y}_{i+}$ for the unmodified peptide. For two conditions $i$ and $i'$ the hypothesis test is formulated as follows:

\begin{equation}
\begin{aligned}
H_{0}: \Delta = ( \mu_{i}^{\ast} - \mu_{i'}^{\ast} ) - ( \mu_{i} - \mu_{i'} ) = 0 \\
H_{a}: \Delta = ( \mu_{i}^{\ast} - \mu_{i'}^{\ast} ) - ( \mu_{i} - \mu_{i'} ) \neq 0
\end{aligned}
\end{equation}

@@

\begin{equation}
\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right]
\end{equation}

@@

\begin{equation}
SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}
\end{equation}

Finally the degrees of freedom are calculated using the following formula.

\begin{equation}
\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right)
\end{equation}

The test statistic for the hypothesis testing is the ratio of the estimate of the adjusted difference to its SE. To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the t distribution with degrees of freedom approximated by the Satterthwaite method\cite{satterthwaite:1946}. Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$\cite{Benjamini:1995}.

\subsubsection*{Design of PTM experiments in the balanced case}
\label{sec:design}

The proposed statistical framework allows for design of PTM experiments in terms of sample size calculation and power analysis. 
%Sample size calculation based on linear models has been described for general applications~\cite{kutner_etal_04a} and specifically for protein significance analysis~\cite{oberg_vitek_09a}. 
Sample size calculation takes as input a) $q$, the desired false discovery rate, b) $\beta$, the average Type II error rate, c) $\Delta$, the minimal log-fold change in adjusted PTM abundance that we would like to detect, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{\gamma^{\ast}}^{2}$ and $\sigma_{\gamma}^{2}$, the anticipated variances associated to modified and unmodified peptide features, respectively. The variances can be derived based on the dataset being analyzed, assuming similar quantitative properties and variations. With these values and a user-specified number of conditions, the corresponding number of technical replicates per condition can then be derived, as described in~\cite{kutner_etal_04a}. Given the above quantities, the minimal number of replicates $J$ is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as
\[
\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2},
\]
where 
\[
\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1},
\]
and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution, with $df = I(J-1)$ degrees of freedom in balanced designs. More details can be found in~\cite{oberg_vitek_09a}.

@@

The statistical modeling approaches discussed above can also be extended to complex designs including experiments with additional sources of variation and unbalanced designs. Equation (1) is the same in this case and the null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.

\begin{align*}
H_{0}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &= 0 \\
H_{a}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &\neq 0
\end{align*}

Model inference is done in the same way to how MSstats~\cite{choi_etal_14a} and MSstatsTMT~\cite{Huang:2020} target the global protein. In cases where the design in unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model. Once the parameters are estimated, we can combine the models using modified versions of the formulas in Section \ref{sec:adjustment_summary}.

The log-fold change in the adjusted PTM abundance, $\Delta$, is now estimated by RML, and Equation (2) becomes 
\[
\hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
\]

The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ and degrees of freedom in Equations (3) and (4) are unchanged from the balanced case, although $\hat{\sigma}_{\gamma^z\ast}$ and $\hat{\sigma}_{\gamma}$ are now estimated using RML.

Details on how the proposed method can be used to run sample size calculations and power analysis on experiments with complex design can be seen in Supplementary Sec. \ref{supp-sec:complex_design}

@@

The code is available on Bioconductor, http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html, and Github, https://github.com/Vitek-Lab/MSstatsPTM.

@@

\subsubsection*{The proposed approach corrects for high FDR in computer simulations}

@@

\subsubsection*{The proposed approach outperforms existing methods in simulations with missing values and low replicates}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In a Label-free benchmark experiment the proposed approach corrects for bias and outperforms existing methods}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{MSstatsPTM corrects for bias in group comparison TMT experiment}

@@

\subsubsection*{Proposed approach removes confounding in time series TMT experiment}

@@

\subsubsection*{Proposed approach is less effective without a separate global profiling run but still fixes bias in Label-free experiment}

@@

\subsubsection*{Power analysis shows more replicates are needed in the PTM profiling run in cases of unequal variance.}

Here we analyze the sample size needed to achieve a desired statistical power. The proposed approach corrects for confounding between the modified peptide and unmodified protein at the cost of increased variation. This can be seen in the calculation for variance in Section \ref{sec:adjustment_summary}. When the variation is increased, the experiment requires a larger number of replicates to reach the same power. Thus the statistical power is dependent on the variance from both the modified peptide and unmodifed protein as well as the number of replicates in each run.
